A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2024-10-10 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: more work on conceptual part

We are looking to find the minimal set of axioms required by a physical theory that allows statistical descriptions. Such a theory must, at the very least, provide us with a notion of ensemble, and a space of ensembles that are allowed in the theory. An ensemble can be understood as the infinite collection of the outputs of a particular preparation procedure. The ensemble space of classical mechanics, for example, is the space of continuous probability distributions of phase space. In quantum mechanics, instead, it is the space of density operators. While the ensemble space for each theory will be in principle different, some basic features will remain the same. We are looking for these basic features, necessary elements that any ensemble space must have.

Note that there may good philosophical grounds to require that a physical theory allows for statistical descriptions. If a physical theory has to be testable, and repeatedly so, then we fail to see how it could be so without providing statistical tools. Every state preparation and measurement, in fact, is typically statistical in nature. The development of such arguments would be interesting to us.

We have three basic requirements on ensembles: first, they have to be identifiable experimentally; second, they need to allow mixtures; lastly, they need an entropy well defined. Let us go through these items one by one.

\subsection{Experimental verifiability}

@@

@@

Suppose we have a physical theory that allows for a statistical description. At the very least, this has to provide a notion of ensemble together with the space of all possible ensembles allowed by the theory. We call this an \textbf{ensemble space} $\Ens$. 
Given a target ensemble $\ens \in \Ens$ and an arbitrary ensemble $\ens[a] \in \Ens$ we will be able to write $\ens = p \ens[a] + (1-p) \ens[b]$ for some $p \in [0,1]$ and some $\ens[b] \in \Ens$. At the very least, we can use $p=0$ and $\ens[b] = \ens$. Given that $p$ is bounded, there will be a biggest $p$ possible which we call the \textbf{fraction} of $\ens[a]$ in $\ens$. Mathematically
\begin{equation}
	\fraction_{\ens}(\ens[a]) = \sup(\{ p \in [0,1] \, | \, \exists \, \ens_1 \in \Ens \text{ s.t. }  \ens = p \ens[a] + \bar{p} \ens_1 \}).
\end{equation}
This quantity tells us how much of ensemble $\ens$ can be constructed, can be characterized, by $\ens[a]$.

>>>>

We are looking to find a minimal set of axioms required by a physical theory that allows statistical descriptions. Such a theory must, at the very least, provide us with a notion of ensemble, and the space of ensembles that are allowed in the theory. An ensemble can be understood as the infinite collection of the outputs of a particular preparation procedure. The ensemble space of classical mechanics, for example, is the space of continuous probability distributions of phase space. In quantum mechanics, instead, it is the space of density operators. While the ensemble space for each theory will be in principle different, some will have many common features. In all cases we must be able to talk about conservative or dissipative processes, measurable properties of the system, probability distributions for outcomes of said properties and so on. We are looking for necessary minimal basic requirements upon which these common tools can be built.

Note that there may good conceptual grounds to require that a physical theory allows for statistical descriptions. If a physical theory is to be repeatedly testable experimentally, then we fail to see how it could be so without providing statistical tools. Every state preparation and measurement, in fact, is typically statistical in nature. Moreover, physical laws are statements of the type ``whenever we prepare this, we can measure that.'' A philosophical exploration of this area would be welcome.

We have identified three basic requirements on ensembles: first, they have to be identifiable experimentally; second, they need to allow mixtures; lastly, they need a well-defined entropy. Let us go through these items one by one.

\subsection{Experimental verifiability}
The first requirement is that ensemble must be connected to experimental verification. That is, we must have enough \textbf{verifiable statements} at our disposal to define ensembles and tell them apart. By verifiable statement we mean a statement for which an experimental test is available that terminates in finite time if and only if the statement is true. For example, the statement ``the mass of the photon is less than $10^{-13} \, eV$'' is verifiable, while ``the mass of the photon is exactly $0 \, eV$'' is not verifiable due to its infinite precision. Mathematically, this requirement forces an ensemble space to be a $T_0$ second countable topological space, where open sets represent verifiable statements and Borel sets represent testable statements without guarantee of termination.

The topological structure, then, represent the most foundational structure for a physical theory. It tells us why functions are ``well-behaved,'' which simply means they have to be topologically continuous in the ``natural topology'' induced by this requirement. It tells why the Banach-Tarski paradox does not apply in physics since non-Borel sets are physically ill-defined as they are not connected to experimental verification. Similarly, it tells us that sets with cardinality greater than that of the continuum are physically not relevant, as they cannot be given a $T_0$ second countable topology.

@@

\subsection{Interplay between structures}

Each structure is important, but the balance between them is even more important. The key is to understand, on physical grounds, where each problem lies and therefore use the appropriate tool. For example, one may focus too much on the convex/linear structure and may be tempted to close the space under infinite convex/linear combinations. Mathematically, it would be very convenient. Physically, it is unsound. It is the link to experimental verification that tells us which limits are allowed, that particles cannot have an infinite average position. Therefore, it is the topology that has to tells us which limits are allowed, not the convex structure. Conceptual clarity has to come first. The math follows. A full examination of these issues is beyond the scope of this article, and we refer to our larger, and future, work. However, there is one key issue that we need to discuss as, we found, it was the main conceptual obstacle to progress.

On the convex structure we can define \textbf{pure states} as the ones that cannot be decomposed into other ensembles. Mathematically, these are the extreme points of the space and all other points can be understood as convex combinations, probability distributions, over them. So, for example, the pure states for a six-sided die are the six possible outcomes, and all ensembles are simply convex combinations $\sum_{i+1}^{6} p_i = 1$. This is exactly how classical discrete spaces work. As physicists, we are somehow trained to think that that's how everything works. In classical mechanics the pure states are perfect preparations of position and momentum. In a classical field theory, they are perfect preparation of all fields at each point in space. Things start breaking down in quantum mechanics, where we do have pure states, but then we also want pure states of position and momentum, which kind of work if you are careful, though we typically aren't. And then there is quantum field theory.

If we look closely, this does not make sense on physical grounds. The fact that the math becomes harder is simply a consequence of trying to fit a square peg in a round hole.

Perfect preparation of classical particles, continuous quantities in quantum mechanics and field configurations in any field theory are not things we can actually achieve. Physically, all ensembles we can prepare have finite precision and finite entropy, and those preparations have neither. They are not ensembles and they are not part of the a convex structure. In the classical case, we can understand them as infinitesimal decompositions: convex combinations that have more and more terms. This is true, but we cannot do that in the quantum case. If we have a pure state, this cannot be decomposed. Therefore an eigenstate of position is not a limit of decomposition, a limit in the convex structure, but a limit of a sequence of pure states, a limit in the topology, and the topology will not even include that limit. We may be able to see a pure state, in some sense, as a superposition of eigenstates of position, but that structure is only available in quantum mechanics, and it does not have a clear physical correspondence.

The insight is that probability distributions for a quantity are not defined by mixing the best ensembles we can prepare (i.e. a convex combination of extreme points), but by assigning a probability over each experimentally verifiable (i.e. open) interval of the possible values of said quantity (its spectra). The two coincide over a discrete set of orthogonal pure states, which is the least physically interesting case. If our are to be physically well defined and true generalizations, they must work in the more complicated case. We not only have to be able to talk about probability in terms of mixtures of ensembles, but we must be able to recover the interval of possible values for physical properties (i.e. the spectra), the object we typically assign probability two.

@@

Note that the mixing coefficient $p$ is connected to probability but it is not itself probability. It represents a parameter for the mixing states, mixing preparations, while probability is typically understood as the chance of something happening, the likelihood of a final event. Mathematically, the probability measure is a function from an event, which can be understood as a collection of fully distinct outcomes, to a number from zero to one. If we are mixing ensemble that represent fully distinct outcomes (i.e. they are orthogonal), then the mixing $p$ is a probability. If we are mixing ensemble that do not represent fully distinct outcomes (e.g. spin up and spin left), the probability of getting one of the outcomes is not the mixing coefficient $p$.

This is exactly what happens in quantum mechanics: probability is defined only on a set of orthogonal states. If we focus only on preparation, then, we only talk about mixing coefficients across non-orthogonal elements and define a more general tool that, when restricted on orthogonal sets, will recover probability.

We concentrate on the following question: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subset \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, if $\ens$ is a uniform distribution over a fair six-sided die, and $A$ contains four ensemble, each fully on ${1,2}$ and uniform distribution over ${5,6}$ two and four, than $5/6$ of $\ens$ can be constructed from $A$, which is the probability that we d

Given a target ensemble $\ens \in \Ens$ and an arbitrary ensemble $\ens[a] \in \Ens$ we will be able to write $\ens = p \ens[a] + (1-p) \ens[b]$ for some $p \in [0,1]$ and some $\ens[b] \in \Ens$. At the very least, we can use $p=0$ and $\ens[b] = \ens$. Given that $p$ is bounded, there will be a biggest $p$ possible which we call the \textbf{fraction} of $\ens[a]$ in $\ens$. Mathematically
\begin{equation}
	\fraction_{\ens}(\ens[a]) = \sup(\{ p \in [0,1] \, | \, \exists \, \ens_1 \in \Ens \text{ s.t. }  \ens = p \ens[a] + \bar{p} \ens_1 \}).
\end{equation}
This quantity tells us how much of ensemble $\ens$ can be constructed, can be characterized, by $\ens[a]$.
